---
output: github_document
---

<!-- README.md is generated from README.Rmd with `make readme`. Please edit that file. -->

```{r setup, include = FALSE}
library(duckknit)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")

# Show duckknit's ``` duckdb fences as ``` sql so GitHub highlights them.
knitr::knit_hooks$set(document = function(x) {
  gsub("(?m)^``` duckdb", "``` sql", x, perl = TRUE)
})
knitr::knit_engines$set(sql = knitr::knit_engines$get("duckdb"))

# The SQL chunks run in the DuckDB CLI with the DuckHTS build bundled in Rduckhts, the
# same extension the page loads in duckdb-wasm. duckknit_start_session() takes no CLI
# flags, so start the process with -unsigned and register it as duckknit's session.
if (!requireNamespace("Rduckhts", quietly = TRUE)) {
  stop("Install Rduckhts before rendering README.md")
}
extension_path <- system.file(
  "duckhts_extension", "build", "duckhts.duckdb_extension",
  package = "Rduckhts", mustWork = TRUE
)
local({
  ns <- asNamespace("duckknit")
  proc <- processx::process$new(
    command = duckknit:::find_duckdb(),
    args = c("-unsigned", ":memory:"),
    stdin = "|", stdout = "|", stderr = "|",
    cleanup = TRUE, cleanup_tree = TRUE
  )
  unlockBinding(".sessions", ns)
  ns$.sessions[["readme"]] <- proc
  lockBinding(".sessions", ns)
  unlockBinding(".state", ns)
  ns$.state$last_session <- "readme"
  lockBinding(".state", ns)

  load_sql <- sprintf(
    "SET allow_extensions_metadata_mismatch=true;\nLOAD '%s';",
    gsub("'", "''", extension_path, fixed = TRUE)
  )
  loaded <- duckknit:::duckknit_exec(proc, load_sql)
  if (nzchar(trimws(loaded$stderr))) stop(loaded$stderr, call. = FALSE)
})
```

# duckpeakwhere <img src="man/figures/logo.svg" align="right" height="139" alt="duckpeakwhere hex sticker" />

[![test](https://github.com/sounkou-bioinfo/duckpeakwhere/actions/workflows/test.yml/badge.svg)](https://github.com/sounkou-bioinfo/duckpeakwhere/actions/workflows/test.yml)

**Where do your peaks land?** Split ChIP-seq, CUT&RUN and ATAC-seq peaks into promoter,
5′ UTR, 3′ UTR, exon, intron and intergenic space, with a genome background bar, in the
browser.

This is a port of [peakwhere](https://github.com/seandavi/peakwhere) by Sean Davis. It
keeps peakwhere's biological rules and its hand-worked test fixture, but replaces its
hand-written GFF3/GTF parser and planned overlap engine with SQL over
[DuckHTS](https://github.com/RGenomicsETL/duckhts) running in
[duckdb-wasm](https://github.com/duckdb/duckdb-wasm).

## How it works

Everything is in [`src/annotate.js`](src/annotate.js), one SQL statement per step:

| Step | peakwhere | here |
|---|---|---|
| Read GFF3 / GTF / BED | hand-written streaming parser | `read_gff`, `read_gtf`, `read_bed` (htslib) |
| Chromosome names (`chr1` = `1`, `M` = `MT`) | JavaScript | `duckhts_contig_key` |
| Transcripts, UTRs, promoter windows | JavaScript model | plain SQL over the feature table |
| Priority partition | planned sorted-segment engine | cut points in SQL; each segment's category from a DuckHTS cgranges index |
| Peak centres, base pairs, genome background | binary search per peak | `duckhts_cgranges_overlaps_list` against the partition |

The rules are peakwhere's:

- Promoter is ±1 kb around the TSS base, counted on the strand.
- Priority is Promoter > 5′ UTR > 3′ UTR > Exon > Intron > Intergenic.
- One count per peak centre, or per base pair.
- Unmatched chromosomes are reported, and a file with more than 5% unmatched peaks is
  not drawn.
- The genome background comes from GFF3 `##sequence-region` lengths.

The page makes no requests outside its own origin. duckdb-wasm, Observable Plot and the
signed DuckHTS builds are served from `vendor/`.

## The same SQL, outside the browser

Nothing in the SQL is browser-specific. These chunks run in the DuckDB CLI against the
bundled example data, with the same DuckHTS extension.

The files are read whole, so the readers use `scan_mode := 'sequential'`, which streams
without looking for a tabix index. Here is the annotation as `read_gff` sees it:

```{sql}
SELECT feature, count(*) AS n
FROM read_gff('examples/gencode.vM25.basic.chr19.gff3.gz', scan_mode := 'sequential')
GROUP BY feature
ORDER BY n DESC, feature
LIMIT 6;
```

The peak files, read by `read_bed` straight from gzip:

```{sql}
SELECT 'CTCF' AS mark, count(*) AS peaks, median("end" - start) AS median_width
FROM read_bed('examples/thymus_CTCF_ENCFF714WDP.chr19.narrowPeak.gz', scan_mode := 'sequential')
UNION ALL
SELECT 'H3K4me3', count(*), median("end" - start)
FROM read_bed('examples/thymus_H3K4me3_ENCFF674JZY.chr19.narrowPeak.gz', scan_mode := 'sequential');
```

And the chromosome names the two files use are matched by key, not by string:

```{sql}
SELECT duckhts_contig_key('chr19') = duckhts_contig_key('19') AS same_contig,
       duckhts_contig_key('chrM') = duckhts_contig_key('MT') AS same_mito;
```

## Results on the example

Peak-centre counts for the mouse thymus chr19 example, from the independent R oracle
[`test/oracle.R`](test/oracle.R). The app's end-to-end test requires the page to
reproduce every number:

```{r thymus, echo = FALSE}
expected <- jsonlite::fromJSON("test/fixtures/thymus-expected.json")$counts_centre
counts <- do.call(rbind, lapply(expected, unlist))
knitr::kable(counts)
```

## Performance

Local W1 (7,220 thymus chr19 peaks) analysis totals in seconds, **median [min–max]**
of five runs after one warm-up. Native replays the app's exact SQL over the same
local HTTP inputs; ChIPseeker uses one-base peak centres but has different annotation
rules, so these are not equal-output speedups. Startup-inclusive times, correctness
gates, category differences, environment and reproduction commands are in the
[performance report](benchmarks/performance.md).

```{r performance, echo = FALSE}
source("bench/report.R")
performance <- jsonlite::read_json("benchmarks/results.json", simplifyVector = FALSE)
knitr::kable(headline(performance), row.names = FALSE)
```

## Limitations

- **Bundled files only.** DuckHTS readers open paths through htslib, which can't see
  files registered with duckdb-wasm (`registerFileBuffer`, `registerFileHandle`) and
  has no handler for `blob:` URLs. So a file dropped on the page can't be read yet. The
  demo reads its example files over same-origin HTTP instead. A test in
  `test/annotate.test.js` records this, and will fail once it changes. Tracked in
  [duckhts#246](https://github.com/RGenomicsETL/duckhts/issues/246).
- **DuckDB 1.5 runtime required.** DuckHTS 1.5.2 fails to load on the current stable
  duckdb-wasm (1.32.0, DuckDB v1.4.3), because it registers SQL that calls
  `read_hts_header` before registering that function
  ([duckhts#247](https://github.com/RGenomicsETL/duckhts/issues/247)). So this pins
  `@duckdb/duckdb-wasm` 1.33.1-dev57.0 (DuckDB v1.5.4).
- Not ported from peakwhere: chrom.sizes input, CSV/TSV peak input with its 1-based
  toggle, and editable bar labels.

## Run it

```bash
npm ci
npm run stage    # download the signed DuckHTS wasm builds, checked against duckhts-manifest.json
npm run vendor   # copy duckdb-wasm and Plot from node_modules into vendor/
npm run serve    # http://127.0.0.1:8000/
npm test         # headless Chromium, needs a Playwright Chromium install
```

`npm run stage` is the only step that uses the network. Once the `duckhts` npm package
is published ([duckhts#248](https://github.com/RGenomicsETL/duckhts/pull/248)), it will
replace both the staging script and `src/duckhts-loader.js`.

## Tests

- `test/annotate.test.js` checks the SQL pipeline against peakwhere's
  [`expected.json`](test/fixtures/expected.json). Those answers were worked out by hand
  before any peakwhere code existed. It covers GFF3 and GTF, peak centres and base
  pairs, the genome background, and `chr`-less names.
- `test/app.test.js` checks the page end to end: the fixture, and the thymus example
  against [`thymus-expected.json`](test/fixtures/thymus-expected.json) from
  [`test/oracle.R`](test/oracle.R). The oracle is a slow brute-force implementation in
  base R that shares no code with the SQL.
- Both check that the signed DuckHTS build loads with unsigned extensions disallowed,
  and that no request leaves the origin.
- CI reruns the oracle and fails if its output differs from the committed expectations.

House rules for contributors and agents are in [`AGENTS.md`](AGENTS.md). Example data
comes from GENCODE and ENCODE; see [`examples/README.md`](examples/README.md).

MIT licensed. See [`LICENSE`](LICENSE).
